---
title: "STAT385 FA19 Final Project"
author: "Eric Wang (wcwang2), Qiqing Xu (qiqingx2), Felix Zhu (zz6)"
date: "12/18/2019"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## STAT 385 Final Project 

### Introduction 

The project uses the `CCSO Bookings` (Champaign County Sheriff's Office Bookings), which contained `67764` observations and `12` columns, and revealed general trend of bookings from `2011` to `2016`. In particular, the project focused on the following variables: 

1. `BOOKING.DATE`: The date of booking in `MM/DD/YYYY` format 
2. `EMPLOYMENT.STATUS`: The employment status of each individual, which included `Student`, `Employed`, `Unemployed` and `Retired` 
3. `RACE`: Race of each individual, which included `Black`, `White`, `Hispanic`, `Asians`, `Others` 
4. `SEX`: Sex of each individual, which included `Male` and `Female` 
5. `MARITAL.STATUS`: Marital status of each individual included `Single`, `Married`, `Divorced`, `Others` 
6. `AGE.AT.ARREST`: The age at arrest of each individual 

### Import and preprocess data 

```{r}
library(shiny) 
data = read.csv('https://uofi.box.com/shared/static/9elozjsg99bgcb7gb546wlfr3r2gc9b7.csv') 
data$BOOKING.DATE = as.Date(data$BOOKING.DATE, format = '%m/%d/%Y') 
data = data[data$Age.at.Arrest <= 100, ] 
data = data[data$EMPLOYMENT.STATUS != '', ] 
```

```{r}
quarter = function(date) { 
  y = substr(as.character(date), 1, 4) 
  q = ceiling(as.integer(substr(as.character(date), 6, 7)) / 3) 
  paste(paste(y, 'Q', sep = ''), as.character(q), sep = '')  
} 
data$quarter = unlist(lapply(data$BOOKING.DATE, FUN = quarter)) 

raceFilter = function(race) { 
  if (race == 'Asian/Pacific Islander') { 
    'Asian'
  } 
  else if (race == 'Black' | race == 'White' | race == 'Hispanic') { 
    as.character(race)
  } 
  else { 
    'Others'
  } 
} 
data$RACE = unlist(lapply(data$RACE, FUN = raceFilter))

maritalFilter = function(marital) { 
  if (marital == 'Single' | marital == 'Married' | marital == 'Divorced') { 
    as.character(marital) 
  } 
  else { 
    'Others' 
  } 
} 
data$MARITIAL.STATUS = unlist(lapply(data$MARITIAL.STATUS, FUN = maritalFilter))
```

### Section 1: Booking date vs. Age at arrest 

An interactive scatterplot in which users can limit the range of dates to see the data for specific dates. The x-axis is the booking date, and the y-axis is the age at arrest. Each observation is colored with a specific color which corresponds to his/her employment status. `Students` are marked orange, `Unemployed` are marked red, `Employed` are marked blue, and `Retired` are marked green. 

```{r}
ui = fluidPage(
  sliderInput("range", "Date of arrest", 
              min = data$BOOKING.DATE[which.min(data$BOOKING.DATE)], 
              max = data$BOOKING.DATE[which.max(data$BOOKING.DATE)], 
              value = c(data$BOOKING.DATE[which.min(data$BOOKING.DATE)], 
                        data$BOOKING.DATE[which.max(data$BOOKING.DATE)]), 
              timeFormat = '%m/%d/%Y'), 
  checkboxInput('employed', 'Employed', TRUE), 
  checkboxInput('unemployed', 'Unemployed', TRUE), 
  checkboxInput('student', 'Student', TRUE), 
  checkboxInput('retired', 'Retired', TRUE), 
  plotOutput("range", height = "500px"),
) 

server <- function(input, output) {
  output$range = renderPlot({ 
    d = data[data$BOOKING.DATE >= input$range[1] & data$BOOKING.DATE <= input$range[2], ] 
    if (!input$student) { 
      d = d[d$EMPLOYMENT.STATUS != 'Student', ] 
    } 
    if (!input$retired) { 
      d = d[d$EMPLOYMENT.STATUS != 'Retired', ] 
    } 
    if (!input$employed) { 
      d = d[d$EMPLOYMENT.STATUS != 'Employed - Full Time' & d$EMPLOYMENT.STATUS != 'Employed - Part Time' & d$EMPLOYMENT.STATUS != 'Self Employed', ] 
    } 
    if (!input$unemployed) { 
      d = d[d$EMPLOYMENT.STATUS != 'Unemployed' & d$EMPLOYMENT.STATUS != 'Laid Off', ] 
    } 
    x = d$BOOKING.DATE
    y = d$Age.at.Arrest 
    df = data.frame(x = x, y = y) 
    df$c = 'blue' 
    df$c[d$EMPLOYMENT.STATUS == 'Retired'] = 'green' 
    df$c[d$EMPLOYMENT.STATUS == 'Student'] = 'orange' 
    df$c[d$EMPLOYMENT.STATUS == 'Unemployed' | d$EMPLOYMENT.STATUS == "Laid Off" ] = 'red' 
    plot(df$x, df$y, col = df$c, xlab="Booking date", ylab="Age at arrest", pch=16, main="Booking date vs. Age at arrest")
  })
}

shinyApp(ui = ui, server = server, options = list(height = 800)) 
```

### Section 2: Booking date vs. Frequency 

An interactive stacked barplot which included the same filters as the first plot. Users could choose among four options: `General`, `Sex`, `Race` and `Marital Status`: 

```{r}
ui = fluidPage(
  checkboxInput('employed', 'Employed', TRUE), 
  checkboxInput('unemployed', 'Unemployed', TRUE), 
  checkboxInput('student', 'Student', TRUE), 
  checkboxInput('retired', 'Retired', TRUE), 
  sliderInput("range", "Date of arrest", 
              min = data$BOOKING.DATE[which.min(data$BOOKING.DATE)], 
              max = data$BOOKING.DATE[which.max(data$BOOKING.DATE)], 
              value = c(data$BOOKING.DATE[which.min(data$BOOKING.DATE)], 
                        data$BOOKING.DATE[which.max(data$BOOKING.DATE)]), 
              timeFormat = '%m/%d/%Y'), 
  selectInput("option", "Select a catagory",
              c("General", "Sex", "Race", "Marital Status")),
  plotOutput("option", height = "600px")
) 

server <- function(input, output) {
  output$option = renderPlot({
    d = data[data$BOOKING.DATE >= input$range[1] & 
               data$BOOKING.DATE <= input$range[2], ] 
    if (!input$student) { 
      d = d[d$EMPLOYMENT.STATUS != 'Student', ] 
    } 
    if (!input$retired) { 
      d = d[d$EMPLOYMENT.STATUS != 'Retired', ] 
    } 
    if (!input$employed) { 
      d = d[d$EMPLOYMENT.STATUS != 'Employed - Full Time' & 
              d$EMPLOYMENT.STATUS != 'Employed - Part Time' & 
              d$EMPLOYMENT.STATUS != 'Self Employed', ] 
    } 
    if (!input$unemployed) { 
      d = d[d$EMPLOYMENT.STATUS != 'Unemployed' & 
              d$EMPLOYMENT.STATUS != 'Laid Off', ] 
    } 
    if (input$option == 'General') { 
      t = table(d$quarter) 
      barplot(t, las=2, col = rainbow(length(t))) 
    } 
    else if (input$option == 'Race') { 
      t = t(as.matrix(table(d$quarter, d$RACE))) 
      t = t[c(2, 5, 3, 1, 4), ] 
      barplot(t, las=2, legend = rownames(t), col = c('blue', 'brown', 'red', 'orange', 'green'))
    } 
    else if (input$option == 'Sex') { 
      t = t(as.matrix(table(d$quarter, d$SEX)))
      t = t[c(3, 2, 1), ] 
      barplot(t, las=2, legend = rownames(t)[c(1, 2)], col = c('blue', 'red'))
    } 
    else { 
      t = t(as.matrix(table(d$quarter, d$MARITIAL.STATUS))) 
      t = t[c(4, 2, 1, 3), ] 
      barplot(t, las=2, legend = rownames(t), col = c('blue', 'orange', 'red', 'green'))
    }
  })
}

shinyApp(ui = ui, server = server, options = list(height = 1000)) 
```

### Section 3: Booking date vs. Probability 

Similar interactive stacked barplot with data normalized so that each quarter data sums to `1`, so that the results are not distorted by the total number of observations during each quarter. 

```{r}
ui = fluidPage(
  checkboxInput('employed', 'Employed', TRUE), 
  checkboxInput('unemployed', 'Unemployed', TRUE), 
  checkboxInput('student', 'Student', TRUE), 
  checkboxInput('retired', 'Retired', TRUE), 
  sliderInput("range", "Date of arrest", 
              min = data$BOOKING.DATE[which.min(data$BOOKING.DATE)], 
              max = data$BOOKING.DATE[which.max(data$BOOKING.DATE)], 
              value = c(data$BOOKING.DATE[which.min(data$BOOKING.DATE)], 
                        data$BOOKING.DATE[which.max(data$BOOKING.DATE)]), 
              timeFormat = '%m/%d/%Y'), 
  selectInput("option", "Select a catagory",
              c("General", "Sex", "Race", "Marital Status")),
  plotOutput("option", height = "600px")
) 

server <- function(input, output) {
  output$option = renderPlot({
    d = data[data$BOOKING.DATE >= input$range[1] & 
               data$BOOKING.DATE <= input$range[2], ] 
    if (!input$student) { 
      d = d[d$EMPLOYMENT.STATUS != 'Student', ] 
    } 
    if (!input$retired) { 
      d = d[d$EMPLOYMENT.STATUS != 'Retired', ] 
    } 
    if (!input$employed) { 
      d = d[d$EMPLOYMENT.STATUS != 'Employed - Full Time' & 
              d$EMPLOYMENT.STATUS != 'Employed - Part Time' & 
              d$EMPLOYMENT.STATUS != 'Self Employed', ] 
    } 
    if (!input$unemployed) { 
      d = d[d$EMPLOYMENT.STATUS != 'Unemployed' & 
              d$EMPLOYMENT.STATUS != 'Laid Off', ] 
    } 
    if (input$option == 'General') { 
      t = table(d$quarter) 
      barplot(t, las=2, col = rainbow(length(t))) 
    } 
    else if (input$option == 'Race') { 
      t = t(as.matrix(table(d$quarter, d$RACE))) 
      t = t[c(2, 5, 3, 1, 4), ] 
      t = sweep(t, 2, colSums(t), FUN = '/') 
      barplot(t, las=2, legend = rownames(t), col = c('blue', 'brown', 'red', 'orange', 'green'))
    } 
    else if (input$option == 'Sex') { 
      t = t(as.matrix(table(d$quarter, d$SEX)))
      t = t[c(3, 2, 1), ] 
      t = sweep(t, 2, colSums(t), FUN = '/') 
      barplot(t, las=2, legend = rownames(t)[c(1, 2)], col = c('blue', 'red'))
    } 
    else { 
      t = t(as.matrix(table(d$quarter, d$MARITIAL.STATUS))) 
      t = t[c(4, 2, 1, 3), ] 
      t = sweep(t, 2, colSums(t), FUN = '/') 
      barplot(t, las=2, legend = rownames(t), col = c('blue', 'orange', 'red', 'green'))
    }
  })
}

shinyApp(ui = ui, server = server, options = list(height = 1000)) 
```